Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-04 paper_jasa.tex
removed signal strength and added visual p-value disccusion

With multiple observers it is possible to estimate the visual $p$-value, and a measure of plot signal strength. The visual $p$-value is calculated from the binomial formula (Equation \ref{binom}). Assuming $H_o$ is true, the probability of picking the actual data plot by chance is $1/m$, and the visual $p$-value is the probability that $x$ or more observers select that plot. Figure \ref{fig:pval_definition} displays this visual $p$-value for each lineup against the conventional $p$value.

Similary, the strength of the signal, is calculated using the proportion of observers picking actual data plot. It is closely related to the power of the test. From Equation \ref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
\[
\hat p_D= (1- {u}/{K})/({m-1})
\]
where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values, on a square root scale. For experiment 3, technically, there are no conventional $p$-values, but the $p$-values are computed in the same manner as experiment 2, assuming that the data satifies the assumptions. The blue line is a loess fit to the data, and the grey line indicates equality of the two axes. The relationship between the conventional $p$-value and the plot signal strength is very strong, particularly when the values are small. As the conventional $p$-value increases, the estimated signal strength increases -- observers find it harder to detect the actual data plot at larger $p$-values. The rise is very sharp for experiment 1, which indicates that observers have a harder time detecting the actual data plot for even small $p$-values.

\green{Where is Figure \ref{fig:pval_definition} referred to and discussed?}
% \blue{Are we going to keep both figures \ref{fig:pval_definition} and \ref{fig:pval_plot_signal}?}
%
%\green{Explain why the term ``plot signal'' is introduced}

>>>>

From the experimental data the visual $p$-values are computed using Definition \ref{dfn:pvalue}. Figure \ref{fig:pval_definition} displays this visual $p$-value for each lineup against the conventional $p$-value.  The pattern of visual $p$-value is interesting. If there is a signal in the data plot, visual $p$-value has a tendency to take a smaller value (small enough to reject $H_0$) and if not it takes a larger value (large enough to not reject $H_0$) giving a clear indication to reject $H_0$ or not. This is why we do not see lot of visual $p$-values in between 0.05 and 0.8 especially for experiment 2. This guide the researcher to make decision confidently while conventional $p$-values take some marginal $p$-values for which researchers get confused whether to reject or 'not reject' $H_0$. For visual test this is not common.

We also see the same pattern from the loess smoother fitted through the points in Figure \ref{fig:pval_definition}. The sigmoid shaped loess curves indicate a clear separation of visual $p$-values. For experiment 1, most of the visual $p$-values are above 0.4 when conventional $p$-values are above 0.2. This is more evident for experiment 2 where most of the visual $p$-values are above 0.8. For experiment 3, we see all the visual $p$ values are almost zero no matter what the conventional $p$-values are. This is expected since conventional test fails to reject $H_0$ where we should reject and visual test does the job nicely.

%With multiple observers it is possible to estimate the visual $p$-value, and a measure of plot signal strength. The visual $p$-value is calculated from the binomial formula (Equation \ref{binom}). Assuming $H_o$ is true, the probability of picking the actual data plot by chance is $1/m$, and the visual $p$-value is the probability that $x$ or more observers select that plot. Figure \ref{fig:pval_definition} displays this visual $p$-value for each lineup against the conventional $p$value.

%Similary, the strength of the signal, is calculated using the proportion of observers picking actual data plot. It is closely related to the power of the test. From Equation \ref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
%\[
%\hat p_D= (1- {u}/{K})/({m-1})
%\]
%where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values, on a square root scale. For experiment 3, technically, there are no conventional $p$-values, but the $p$-values are computed in the same manner as experiment 2, assuming that the data satifies the assumptions. The blue line is a loess fit to the data, and the grey line indicates equality of the two axes. The relationship between the conventional $p$-value and the plot signal strength is very strong, particularly when the values are small. As the conventional $p$-value increases, the estimated signal strength increases -- observers find it harder to detect the actual data plot at larger $p$-values. The rise is very sharp for experiment 1, which indicates that observers have a harder time detecting the actual data plot for even small $p$-values.

%\green{Where is Figure \ref{fig:pval_definition} referred to and discussed?}
% \blue{Are we going to keep both figures \ref{fig:pval_definition} and \ref{fig:pval_plot_signal}?}
%
%\green{Explain why the term ``plot signal'' is introduced}
